---
title: "Formula 1 - Analyze to Action"
subtitle: "Proposal"
author: 
  - name: "Pixel and Patterns"
    affiliations:
      - name: "College of Information Science, University of Arizona"
description: "The project focuses on analyzing and visualizing Formula 1 race data using the f1DataR package, which provides access to historical race results, driver information, and circuit details. The aim is to extract insights from various seasons, focusing on specific tracks and team performance, including visualizing race times, positions, and the influence of circuit characteristics. Through this analysis, the project aims to offer a deeper understanding of race outcomes, driver performance, and trends across different seasons, with a particular focus on specific Grand Prix events."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
library(f1dataR)
library(ggplot2)
library(dplyr)
library(ggmap)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false

library(f1dataR)

#help(package = "f1DataR")

#Example

# Load the drivers data for 2021 season
drivers_data <- load_drivers(2021)
head(drivers_data)
```

```{r}
# Example: load session results for a specific round in 2021
session_results <- load_session_laps(2021, round = 1)
head(session_results)
```

## Questions

1\) **How can we visualize the race winners on the Circuit de Monaco over the past four years (2019-2023) in a clear and insightful way using R?**

### Description:

To visualize the race winners at the Circuit de Monaco (with a specific circuit ID), we can use the `f1dataR` package to load race session data for the past four years. We will filter the data for races held on the Circuit de Monaco Circuit, identify the winners of each race, and plot the results on a race track or through a race track graph.
